---
title: "Introduction to resources and data sets"
author: "Daniel J. Eck"
date: ""
output: beamer_presentation
urlcolor: blue
header-includes: 
 - \usepackage{amsthm}
 - \usepackage{amsmath}
 - \usepackage{amsfonts}
 - \usepackage{amscd}
 - \usepackage{amssymb}
 - \usepackage[sectionbib]{natbib}
 - \usepackage{url}
 - \usepackage{graphicx,times}
 - \usepackage{array,fancyhdr,rotating}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome

Welcome to STAT 430: Baseball Analytics! 

\vspace{12pt}

Let's have a fun semester exploring the ability of statistics to quantify winning games and evaluating players.

\vspace{12pt}

These brief lecture slides are meant to introduce data sets used in class and supplement Chapter 1 in your textbook. They will also go over course logistics.





## GitHub

GitHub is a a cloud-based service that implements a Git repository hosting system.

\vspace{12pt}

Course materials will be distributed from my GitHub organization which was built using CS's GitHub-repo-creator. Your repo can be created through this link
\begin{center}
\url{https://edu.cs.illinois.edu/create-gh-repo/sp25_stat430}
\end{center}

\vspace{12pt}

See the \texttt{setup.md} file in the **stat430sp25** root directory for details on setting up GitHub in this course.


## Software

 - The R Project for Statistical Computing: \url{https://www.r-project.org/}

 - RStudio as an integrated development environment for R: \url{https://www.rstudio.com/}

\vspace{12pt}

Install both R and RStudio and consider creating an [RStudio project](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) for better organization.


## Data sets used in class

- \texttt{Lahman} package: 
```{r, eval = FALSE}
install.packages("Lahman")
```
- [Retrosheet](www.retrosheet.org/gamelogs/index.html) data. Appendix A in your textbook provides an R script file for downloading and parsing all the game log files. A possible more convenient approach for obtaining retrosheets is included in this slide deck. Or you can occasionally use the \texttt{retrosheet} package for simple retrosheets:
```{r, eval = FALSE}
install.packages("retrosheet")
```
<!-- - PITCHf/x data in XML format.  Appendix B in your textbook details the \texttt{xml2} package and other resources for working with XML data in R.  -->


## 

- Statcast data obtained from the \texttt{baseballr} package: 
```{r, eval = FALSE}
install.packages("baseballr")
```
- [baseball_R](https://github.com/beanumber/baseball_R). Coding scripts and data sets that supplement your textbook.
- Statcast data obtained from the \texttt{bbd} package: 
```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("daviddalpiaz/bbd")
```

The \texttt{bbd} package was created by UIUC CS professor [David Dalpiaz](https://daviddalpiaz.org/). [Here is a link](https://github.com/daviddalpiaz/bbd) to the GitHub repo.

\vspace{12pt}

Statcast and Retrosheet data sets will be provided if past difficulties persist.


## 

- Data scraped from [baseball reference](https://www.baseball-reference.com/): 
 
\tiny
```{r, eval = FALSE}
bwar_bat = readr::read_csv("https://www.baseball-reference.com/data/war_daily_bat.txt", 
                           na = "NULL")
bwar_pit = readr::read_csv("https://www.baseball-reference.com/data/war_daily_pitch.txt", 
                           na = "NULL")
```

\normalsize
- Era-adjusted data sets are accessible in the \texttt{fullhouse} R package:
```{r, eval = FALSE}
#install.packages("devtools")
install_github(repo = "DEck13/fullhouse")
```
This is beta version of an R package developed by myself and my students. It contains era-adjusted baseball statistics that appear on [our website](https://eckeraadjustment.web.illinois.edu/). These stats are obtained through an application of what we call a ``Full House Model," a statistical approach presented in a paper that was recently accepted for publication.

## Introduction to Lahman

The \texttt{Lahman} package contains several tables consisting of useful stat. We highlight a few tables below

\vspace{12pt}

```{r}
#install.packages("Lahman") 
library(Lahman)
data(Batting)
data(Pitching)
data(Fielding)
data(Teams)
```


## Lahman Batting table

\tiny
```{r}
head(Batting)
```


## Lahman Pitching table

\tiny
```{r}
head(Pitching)
```


## Lahman Fielding table

\tiny
```{r}
head(Fielding)
```


## Lahman Teams table

\tiny
```{r}
head(Teams, 3)
```


## Retrosheets

There is a lot of box score information contained in a retrosheet. 


Basic retrosheets can be obtained from the \texttt{retrosheet} package (the following code chunk has \texttt{eval = FALSE} because the retrosheet will not fit on a single slide).

\vspace{12pt}

```{r, eval = FALSE}
library(retrosheet)
getRetrosheet(type = "game", year = 2012)
```


##

More comprehensive retrosheets can be obtained from the \texttt{baseballr} package (the following code chunk has \texttt{eval = FALSE} because the retrosheet will take awhile to load and will be stored locally).

\vspace{12pt}

\tiny
```{r, eval = FALSE}
library(baseballr)
retrosheet_data(path_to_directory = "~/Desktop/baseball_course/retrosheet", 
                years_to_acquire = 1998)
```

\normalsize
Obtaining retrosheets via \texttt{baseballr} requires some work outlined by Bill Petti [here](https://billpetti.github.io/2019-08-10-build-retrosheet-event-roster-database-rstats-baseballr/). 

The steps in the hyperlink above require one to first download and install files from [the Chadwick Bureau](https://github.com/chadwickbureau/chadwick/releases).

Follow the instructions in the INSTALL file in the downloaded Chadwick tarball (this course used \texttt{chadwick-0.9.5}).




